Tracking and predicting turn structure during language acquisition

### 2014-09-22 paper/CasillasFrank-Anticip.tex
minor tweaks

\begin{abstract}
We investigate language acquisition through the lens of a fundamental conversational skill: turn taking. Children begin developing turn-taking skills in infancy, but take several years to assimilate language into their turn-taking systems. In two eye-tracking experiments with children ages 1--6, we measured participants' spontaneous predictions about upcoming speaker change while controlling the amount of linguistic information available. We found that children already predicted upcoming turns at age one, and integrated linguistic cues differently at different ages. Children under 3;0 showed a general advantage for prosody over lexicosyntax, contrary to the typical finding for adults. Children 2;0 and older showed more predictive switches for questions than non-questions, but only when lexical information was available. But, we found no evidence that lexicosyntax is sufficient for turn prediction---participants' performance was best overall with access to lexicosyntax and prosody together. Our results point to the importance of interaction in children’s real-time linguistic processing.
\end{abstract}

@@

In another study, \citet*{torreiraUndRev} used a combination of button-press and verbal responses to investigate the relationship between lexicosyntactic and prosodic cues in turn-boundary identification. They found that participants never verbally responded or pressed the ``turn-end'' button when hearing a syntactically-complete unit without a boundary tone. However, utterance-level prosodic boundaries embedded in multi-utterance turns \textit{could} trick participants into anticipating a turn end 29\% of the time (see also de \citet{de-ruiter2006}:525). Their findings corroborate other corpus and experimental findings promoting a combination of cues (lexicosyntactic, prosodic, and pragmatic) as key for accurate turn-end prediction \citep{duncan1972, ford1996, hirvenkari2013}.

@@

There are mixed results regarding when children begin to anticipate turn structure in conversation. One study found that 12-month-olds make more predictive gaze shifts while watching human verbal conversation compared to object non-speech ``conversation'' \citep{bakker2011}, but another only found a similar effect at 36 months \citep{hofsten2009}. However, neither of these two studies had baselines to which the turn-relevant looking behavior could be compared. A baseline measurement is critical because there may be developmental differences in gaze shifting between conversational participants, even if the shifting is not related to turn structure. Such developmental differences could produce artifactual changes in measures of turn-contingent shifting.

@@

Although the Keitel et al. \citeyearpar{keitel2013} study constitutes a substantial advance over previous work, it nevertheless has several limitations. Because these limitations directly inform our own study design, we review them in some detail. First, their estimates of baseline ``random'' gaze frequency were computed from switches that occurred while the speaker's turn was ongoing (i.e, during turns, and not during inter-turn gaps). Their estimates of anticipatory shifting frequency were computed from switches that occurred within the time window immediately surrounding inter-turn gaps (i.e., around inter-turn gaps, and primarily not during turns). Assuming that gaze switches happen most often during the inter-turn gap \citep{hirvenkari2013}, switches made during ongoing speech (their ``random'' baseline frequency) would be at a minimum. Therefore their ``random'' (baseline) frequency actually maximized the possibility of finding a difference between ``random'' and anticipatory gaze frequencies. A more conservative baseline would be to compare participants' looking behavior at turn transitions to their looking behavior during randomly-selected windows of time throughout the stimulus. We follow this latter approach in our work.

@@

Children showed increased anticipatory switching with age across all conditions, making almost as many transitions as adults ages five and six. Adults reacted to the \textit{no speech} condition differently than the children did, tracking the interaction less closely than in the three linguistic conditions. They may have minimized switching in the \textit{no speech} condition because they realized that there was no conversation to follow.

@@

The core aims of Study 2 were to gain better traction on the individual roles of prosody and lexicosyntax in children's turn predictions, and to expand our age range to capture more developmental change. We found that effects of linguistic processing and age primarily surface within the question effect: Children and adults make more anticipatory switches after questions than non-questions, but do so more often when they have greater access to linguistic information, whether increased access came with age or with linguistic condition. Our results hint that lexical information is critical in identifying questions during conversation, especially for children. Even so, performance was consistently best in the \textit{normal} condition, in which participants had lexical and prosodic information together.

By extending the age range of participants, we were able to see more of the developmental trajectory in children's predictions about turn structure. We saw that one-year-olds already made more anticipatory switches than would be expected by chance in two of the four conditions. As they got older, children demonstrated above-chance anticipatory switches in more conditions and made more anticipatory switches overall. Similar to Experiment 1, children between ages three and five behaved similarly to each other, and comparably to adults, showing small but steady gains in their anticipatory switches with age. Though the advantage for questions generally increased with age, the biggest developmental shift in our data occurred between one and two years, when the question effect began to reliably emerge in the two lexical conditions. It will be up to future work to disentangle which lexical (and prosodic) cues children begin to pick out in distinguishing questions from other speech acts, and when.

@@

In two experiments investigating children's anticipatory gaze to upcoming speakers, we found evidence that turn prediction develops early in childhood, and that linguistic cues are integrated differently at different ages. Children under 3;0 showed an advantage for prosody over lexicosyntax, contrary to the typical finding for adults in prior work \citep{de-ruiter2006}. Children 2;0 and older also showed a slight advantage for lexicosyntax in the form of a transition type effect: children made more anticipatory switches following questions than non-questions in conditions where lexical information was present. We found no evidence that, for children or adults, lexicosyntax alone was ``sufficient'' (or otherwise equal) compared to full linguistic information) for spontaneous turn prediction \citep[pg. 531]{de-ruiter2006}: participants' performance was best in conditions when they had access to the full linguistic signal.

@@

Although children showed an initial advantage for prosody, their predictions before 3;0 were probably limited to basic prosodic contours and boundaries, and simple cues such as final lengthening and turn-final silence \citep{pannekamp2006}. More complex structural predictions would have required them to access lexicosyntactic information because higher-level structure in the prosodic signal is linked to the syntactic structure being used. The close link between prosodic and syntactic structure makes it difficult to tease apart how predictive processing in one domain (e.g., prosody) is distinct from the other (e.g., syntax) in turn prediction. We leave this challenge to future work.

@@

Thanks! (To be filled in, but must include NSF GRFP, ERC, Bing, and the CDM).

>>>>

\begin{abstract}
We investigate language acquisition through the lens of a fundamental conversational skill: turn taking. Children begin developing turn-taking skills in infancy, but take several years to assimilate their growing knowledge of language with this ability. In two eye-tracking experiments with children across a wide developmental range, we measured spontaneous predictions about upcoming speaker change while controlling the amount of linguistic information available. We found that children already predicted upcoming turns at age one, but they integrated linguistic cues differently at different ages. Children under three showed a general advantage for prosody over lexicosyntax, contrary to the typical finding for adults. Children two and older showed more predictive switches for questions than non-questions, but only when lexical information was available. We found no evidence that lexicosyntax alone guides turn prediction---instead, participants' performance was best overall with access to lexicosyntax and prosody together. Our results point to the importance of interaction in children's real-time linguistic processing.
\end{abstract}

@@

In another study, \citet*{torreiraUndRev} used a combination of button-press and verbal responses to investigate the relationship between lexicosyntactic and prosodic cues in turn-boundary identification. They found that participants never verbally responded or pressed the ``turn-end'' button when hearing a syntactically-complete unit without a boundary tone. Utterance-level prosodic boundaries embedded in multi-utterance turns \textit{could} trick participants into anticipating a turn end 29\% of the time, however (see also de \citet{de-ruiter2006}:525). Their findings corroborate other corpus and experimental findings promoting a combination of cues (lexicosyntactic, prosodic, and pragmatic) as key for accurate turn-end prediction \citep{duncan1972, ford1996, hirvenkari2013}.

@@

There are mixed results regarding when children begin to anticipate turn structure in conversation. One study found that 12-month-olds make more predictive gaze shifts while watching human verbal conversation compared to object non-speech ``conversation'' \citep{bakker2011}, but another only found a similar effect at 36 months \citep{hofsten2009}. Neither of these two studies had baselines to which the turn-relevant looking behavior could be compared, however. A baseline measurement is critical because there may be developmental differences in gaze shifting between conversational participants, even if the shifting is not related to turn structure. Such developmental differences could produce artifactual changes in measures of turn-contingent shifting.

@@

Although the Keitel et al. \citeyearpar{keitel2013} study constitutes a substantial advance over previous work, it nevertheless has several limitations. Because these limitations directly inform our own study design, we review them in some detail. First, their estimates of baseline gaze frequency (``random'' in their terminology) were computed from switches that occurred specifically while the speaker's turn was ongoing; in contrast, estimates of anticipatory shifting frequency were computed from switches that occurred within the time window immediately surrounding inter-turn gaps. Assuming that gaze switches happen most often during the inter-turn gap \citep{hirvenkari2013}, switches made during ongoing speech (their baseline) would be minimized. This baseline thus maximized the possibility of finding a difference between ``random'' and anticipatory gaze frequencies. A more conservative baseline would be to compare participants' looking behavior at turn transitions to their looking behavior during randomly-selected windows of time throughout the stimulus. We follow this conservative approach in our work.

@@

Children showed increased anticipatory switching with age across all conditions, making almost as many transitions as adults by ages five and six. Adults reacted to the \textit{no speech} condition differently than the children did, tracking the interaction less closely than in the three linguistic conditions. They may have minimized switching in the \textit{no speech} condition because they realized that there was no conversation to follow.

@@

The core aims of Study 2 were to gain better traction on the individual roles of prosody and lexicosyntax in children's turn predictions, and to expand our age range to capture more developmental change. We found that effects of linguistic processing and age were present in this dataset, but both were primarily found within the question effect. Children and adults made more anticipatory switches after questions than non-questions, but did so more often when they had greater access to linguistic information, whether increased access came with age or with linguistic condition. Our results hint that lexical information is critical in identifying questions during conversation, especially for children. Even so, performance was consistently best in the \textit{normal} condition, in which participants had lexical and prosodic information together.

By extending the age range of participants, we were able to see more of the developmental trajectory in children's predictions about turn structure. One-year-olds already made more anticipatory switches than would be expected by chance in two of the four conditions. As they got older, children demonstrated above-chance anticipatory switches in more conditions and made more anticipatory switches overall. Similar to Experiment 1, children between ages three and five behaved similarly to each other, and comparably to adults, showing small but steady gains in their anticipatory switches with age. Though the advantage for questions generally increased with age, the biggest developmental shift in our data occurred between one and two years, when the question effect began to reliably emerge in the two lexical conditions. It will be up to future work to disentangle which lexical (and prosodic) cues children begin to pick out in distinguishing questions from other speech acts, and when.

@@

In two experiments investigating children's anticipatory gaze to upcoming speakers, we found evidence that turn prediction develops early in childhood, and that linguistic cues are integrated differently at different ages. Children under three showed an advantage for prosody over lexicosyntax, contrary to the typical finding for adults in prior work \citep{de-ruiter2006}. Children two and older also showed a slight advantage for lexicosyntax in the form of a transition type effect: children made more anticipatory switches following questions than non-questions in conditions where lexical information was present. We found no evidence that, for children or adults, lexicosyntax alone was ``sufficient'' (equal to full linguistic information) for spontaneous turn prediction (cf. \citealp{de-ruiter2006}): Participants' performance was best in conditions when they had access to the full linguistic signal.

@@

Although children showed an initial advantage for prosody, their predictions before age three were probably limited to basic prosodic contours and boundaries, and simple cues such as final lengthening and turn-final silence \citep{pannekamp2006}. More complex structural predictions would have required them to access lexicosyntactic information because higher-level structure in the prosodic signal is linked to the syntactic structure being used. The close link between prosodic and syntactic structure makes it difficult to tease apart how predictive processing in one domain (e.g., prosody) is distinct from the other (e.g., syntax) in turn prediction. We leave this challenge to future work.

@@

We gratefully acknowledge the parents and children at Bing Nursery School, Children's Discovery Museum of San Jose. This work was supported by an ERC grant XYZ, an NSF GRFP to the first author, and a Merck Foundation Fellowship to the second author. An earlier version of these data and analyses were initially presented to conference audiences in \cite{casillas2012} and \cite{casillas2013}.
